New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2023-01-27 paper/paper.tex
grammarly changes

\begin{abstract}
This paper describes new user controls for examining high-dimensional
data using low-dimensional linear projections and slices. A user can
interactively change the contribution of a given variable to a
low-dimensional projection, which is useful for exploring the
sensitivity of structure to particular variables. The user can also
interactively shift the center of a slice, for example, to explore how
structure changes in local subspaces. The Mathematica package as well as
example notebooks are provided, which contain functions enabling the
user to experiment with these new manual controls, with one specifically
for exploring regions and boundaries produced by classification models.
The advantage of Mathematica is its linear algebra capabilities, and
interactive cursor location controls. Some limited implementation has
also been made available in the R package \tt{tourr}. 
\end{abstract}

@@

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications,
\textcolor{blue}{methods which work for arbitrary dimensions have more general utility}.
A good approach is to show projections from an arbitrary dimensional
space to create dynamic data visualizations called \emph{tours}. Tours
involve views of high-dimensional data with low-dimensional projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. (Various videos of tours, e.g.
\citet{dataviewer}, \citet{gt-pp-video}, on the \citet{ASA22} illustrate
this concept.) Prior to Asimov's work, there were numerous preparatory
developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video \citep{PRIM9-video}
illustrating the capabilities is available through \textcolor{blue}{the}
video library of \citet{ASA22}.) Steering through all possible
projections is impossible, unlike Asimov's tours which allows one to
quickly see many, many different projections. After Asimov there have
been many tour developments, which are summarized in \citet{lee2021}.

One such direction of work develops the ideas from PRIM-9, to provide
manual control of a tour. \citet{cook_manual_1997} describe controls for
1D (or 2D) projections, respectively in a 2D (or 3D) manipulation space,
allowing the user to select any variable axis, and rotate it into, or
out of, or around the projection through horizontal, vertical, oblique,
radial or angular changes in value. \citet{spyrison_spinifex_2020}
refined this algorithm and implemented it to generate radial tour
animation sequences.

Manual controls are especially useful for assessing sensitivity of
structure to particular elements of the projection. There are many
places where it is useful. In exploratory data analysis, where one sees
clusters in a projection, one may ask whether some variables can be
removed from the projection without affecting the clustering. For
interpreting models, one can reduce or increase a variable's
contribution to examine the variable importance. Having the user
interact with a projection is extremely valuable for understanding
high-dimensional data. However, these algorithms have two problems: (1)
the pre-processing of creating a manipulation space overly complicates
the algorithm, (2) extending to higher dimensional control is difficult.

Another potentially useful manual control, is to allow the user to
choose the position of the center of a slice. The slice tour was
introduced in \citet{slicetour}. It operates by converting the
projection plane into a slice, by removing or de-emphasizing points that
are further than a fixed orthogonal distance from the plane. The
projection plane is usually thought of as passing through the center of
the data. Manual control would allow the user to change the position of
the center point, by shifting it along a coordinate axis, while keeping
the orientation of the projection plane fixed. The purpose would be to
explore how or if the shape of the data, in the space orthogonal to the
projection, changes as one gets away from the center. It would also
allow the user to interactively decide on the thickness of the slice.

This paper explains the new manual controls for projection and slice
tours. The next section describes the new algorithm for manual control,
for both projections and slices. The use of these methods is illustrated
to compare and contrast boundaries constructed by different classifiers.
The software section describes a Mathematica \citep{Mathematica} package
that is used for the application, and describes the interactive
environment that would be desirable within R \citep{rref} as new
technology becomes available. The paper is accompanied by an appendix
with more details and adjustments to the manual controls, and three
Mathematica notebooks that can be used to reproduce the applications.

@@

The methods for updating component values in \citet{cook_manual_1997}
(and utilized in \citet{spyrison_spinifex_2020}) are prescribed
primarily for a 2D projection, to take advantage of (then) newly
developed 3D trackball controls made available for computer gaming. The
first step was to construct a 3D manipulation space from a 2D
projection. In this space, the coefficient of the controlled variable
ranges between -1 and 1. Movements of a cursor are recorded and
converted into changes in the values of \(V_m\) thus changing the
displayed 2D projection. The algorithm also provided constraints to
horizontal, vertical, radial or angular motions only. The construction
of the manipulation space overly complicates the manual controls,
especially when considering possible techniques that will apply to
arbitrary \(d\).

@@

The new approach emerged from experiments on the tour using the linear
algebra capabilities, and relatively new interactive graphics interface,
available in Mathematica. The components corresponding to \(V_m\) are
directly controlled by cursor movement, which updates row \(m\) of
\(A\). The updated matrix is then orthonormalised.

@@

This algorithm will produce the changes to a projection as illustrated
in Figure \ref{fig:manualsequence}. The controlled variable, \(V_m\),
corresponds to the dark line, and sequential changes to row \(m\) of
\(A\) can be seen to roughly follow a specified position (dot). Changes
in the other components happen as a result of the orthonormalisation,
but are uncontrolled.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/manualsequence-1} \caption{Sequence of projections where contribution of one variable is controlled (dark) is changed using unconstrained orthonormalisation. The dot indicates the chosen values for the controlled variable, $V_m$. It can be seen that the actual axis does not precisely match the chosen position, but it is close.}\label{fig:manualsequence}
\end{figure}

@@

The problem with the new simple method is that it is not faithful to the
precise values for \(V_m\) because the orthonormalisation will change
them. Even though these changes are for the most part imperceptible, one
may wish to avoid them and there are numerous ways that this can be
enforced, a few are detailed in the Appendix. These primarily differ in
how the remaining variables are adjusted during orthonormalisation.

@@

To better explore the space we combine the manual controls for the
projection with manual controls for slicing. A slice is a section of the
data that is defined by a projection, a center point that is anchoring
it in the high-dimensional space and the slice thickness \(h\)
\citep{slicetour}. A data point is inside the slice if its orthogonal
distance from the projection plane (passing through the center point) is
below the thickness \(h\). This orthogonal distance is computed in terms
of the component that is normal on the projection plane. For
\(\mathbf{x}_i\) a \(p\) dimensional data point and \(\mathbf{c}\) the
center point (in the same \(p\) dimensional space) we compute the
orthogonal distance as \begin{equation}
v_i^2 = ||\mathbf{x}_i' - \mathbf{c}'||^2 = \mathbf{x}_i'^2 + \mathbf{c}'^2 - 2 \mathbf{x}_i'\cdot\mathbf{c}' ,
\label{eq:slice}
\end{equation} with
\(\mathbf{c}' = \mathbf{c} - (\mathbf{c}\cdot \mathbf{a}_1) \mathbf{a}_1 - (\mathbf{c}\cdot \mathbf{a}_2 )\mathbf{a}_2\),
\(\mathbf{x}_i' = \mathbf{x}_i - (\mathbf{x}_i\cdot \mathbf{a}_1) \mathbf{a}_1 - (\mathbf{x}_i\cdot \mathbf{a}_2) \mathbf{a}_2\)
and \(\mathbf{a}_k, k=1,2 (=d)\) denoting the columns of the projection
matrix, \(\mathbf{A}=(\mathbf{a}_1, \mathbf{a}_2)\).

@@

In addition it is also useful to interactively change the slice
thickness \(h\) (also called the slice radius), in particular to find
the preferred value for exploring the input data. For guidance the
estimates of the number of points inside the slice as a function of the
original sample size \(N\) and the number of dimensions \(p\) from
\citet{sectionpursuit} can be used: in case of a uniform distribution
inside a sphere of radius \(R\) a slice with thickness \(h\) will
contain \(N_S\) points, with \begin{equation}
N_S(h, p, R, N) = \frac{N}{2} \left(\frac{h}{R}\right)^{p-2} \left(p - (p-2)\left(\frac{h}{R}\right)^{2}\right).
\label{eq:count}
\end{equation}

@@

In addition, we also want to be able to look at slices of the data and
select slicing parameters interactively. Here we have implemented a
switch to change between projection and slice view, a slider that
adjusts the slice thickness, \(h\) (named \texttt{height} in the
Mathematica interface), and a numeric input to specify the center point
\(c\) explicitly. The display will update based on these input values,
in particular the view will jump to a new slice when \(c\) is changed.

\textcolor{blue}{Within R, the natural implementation for an interactive interface would use}
\texttt{shiny} \citep{shinym} with \texttt{plotly} \citep{plotly}
drawing. However, the \texttt{plotly} package is designed to add on to
\texttt{ggplot2} graphics, not to be a programmable interactive graphics
system. It lacks capability to implement the primary tool needed for a
manual tour, that is the tracking of the cursor position in small
increments. In theory, the historical \texttt{locator()} function of
base R should provide the necessary capability, but it's behavior is not
consistent between graphical devices. The cursor position control is
possible with \texttt{rgl} \citep{rgl} but this package is focused on
3D, not the arbitrary dimension of tours. The level of interactivity
would have been possible with the \texttt{cranvas} system
\citep{cranvas} or \texttt{rJava} \citep{rJava} but neither has a
graphics toolkit that currently operates across platforms. What we have
chosen to do is to use the interactivity of Mathematica to experiment,
and to implement as an animation in the \texttt{tourr} \citep{tourr}
package. There is hope that better interactivity will be available for R
soon, perhaps in javascript \citep[e.g.][\citet{detourr}]{langevitour},
which might allow for the experimentation to develop new techniques.

@@

Changing slice center manually can be accomplished with the new
function, \texttt{manual\_slice()}. This changes the value of the center
point (\(\mathbf{c}\)) of the slice, along a selected variable axis. It
slices from the center of the data out to one side, back to the center,
out to the opposite side, and then back to the center.

@@

\caption{Scatterplot matrix of the (standardised) penguins data. The three species are reasonably different in size, with Gentoo distinguised from the other two on body depth relative to flipper length and body mass.}\label{fig:penguins-scatmat}
\end{figure}

To illustrate the usefulness of the manual controls we use the 4D
penguins data \citep{penguins} and look at classification models
following \citet{sam11271}. We will show how classification boundaries
can be explored and better understood on projections and slices through
4D space. Figure \ref{fig:penguins-scatmat} shows a scatterplot matrix
of this data. There are four variables
(\texttt{bl\ =\ bill\_length\_mm,\ bd\ =\ bill\_depth\_mm,\ fl\ =\ flipper\_length\_mm,\ bm\ =\ body\_mass\_g})
measuring the size of the penguins from three species (Adelie, Chinstrap
and Gentoo). The scatterplot matrix shows that the three species appear
to be likely separable, and that at least the Gentoo can be
distinguished from the other two species when \texttt{bd} is paired with
\texttt{fl} or \texttt{bm}. The steps for exploring boundaries in this
example are as follows:

@@

We start by exploring the projections of the model prediction. Figure
\ref{proj1} summarizes the process. Rotating manually the view, we can
visualize the location of predictions for each of the three species, and
also get a sense of the difference between the two models. To illustrate
this difference, we have manually rotated the projection for the RF
model (left plot) to identify a view that shows the non-linear but
block-type structure that is typical of this type of model. This
particular projection (\(A_1\)) is exported so that it can later be used
to show the LDA model (middle plot) and the actual data (right plot).
What can be seen is the linearity of the LDA model, where the boundaries
are linear and oblique to the variable axes. And, interestingly, this
particular projection of the original data shows very distinct clusters
of the three species. That means, the obscuring of the boundaries
between groups for both of the models is driven by what is happening in
the orthogonal space to the plane of the selected projection.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.95\textwidth]{figures/proj1.pdf}}
%\centerline{\includegraphics[width=0.32\textwidth]{figures/proj1_rf.png}
%\includegraphics[width=0.32\textwidth]{figures/proj1_lda.png}
%\includegraphics[width=0.32\textwidth]{figures/proj1_data.png}}
\caption{A projection identified using the manual tour, because it shows differences in classification boundaries induced by different models. A RF model produces blocky, rectangular, boundaries, while the LDA model produces oblique linear boundaries. For comparison, in this projection the data shows the three groups are almost perfectly separable.}
\label{proj1}
\end{figure*}

This example can be reproduced with the code in
penguins\_exploring\_manually.nb and a run through is shown in the video
at \url{https://vimeo.com/747585410}.

@@

We now continue the investigation by slicing orthogonally to the
projection \(A_1\). For both models we look at a thin (\(h=0.5\)) slice
through the center, \(\mathbf{c}^0 = (0,0,0,0)\). At first we explore
how changing the projection away from \(A_1\) can help with
understanding the boundary better. For our example, notice that \(A_1\)
does not contain any contribution from the second variable
(\texttt{bd}), so we will first rotate this variable into the view.

@@

The RF is almost straight in this view, as something we might expect
from a tree model where splitting occurs on single variables only.
However, it is straight in a combination of the second and third
variable (\texttt{bd} and \texttt{fl}). By examining the scatterplot
matrix, we can understand how this boundary was built: on each of
\texttt{bd} and \texttt{fl} it is possible to cut on a value where most
of the Gentoo penguins are different from the other two species for any
tree, and likely only one is needed. Thus the forest construction is
providing a sample of trees that use one variable or the other variable
to split, producing the blocky boundary on the combination of the two.

In contrast, the LDA boundary has an oblique split between the two, and
roughly divides the space into three similarly sized areas. It is almost
like a textbook illustration of how LDA works for two variables (2D)
where by assuming the clusters have equal variance-covariance, it places
a boundary half-way between the group means.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.95\textwidth]{figures/proj2.pdf}}
%\centerline{\includegraphics[width=0.45\textwidth]{figures/slice1_rf.png}
%\includegraphics[width=0.45\textwidth]{figures/slice1_lda.png}}
%\centerline{\includegraphics[width=0.45\textwidth]{figures/slice2_rf.png}
%\includegraphics[width=0.45\textwidth]{figures/slice2_lda.png}}
\caption{Comparing slices from two different projections $A_1$ (top row) and $A_2$ (bottom row), to compare and contrast classification boundaries induced by two models RF and LDA. A thin slice through the center of the data is shown. The plot of the data shows that the Gentoo group (light grey/green) is less separated in $A_1$, and the rotation to $A_2$ produces a bigger gap. The boundaries between groups are clear in $A_2$ also. The LDA boundary is more oblique between Gentoo and Chinstrap than the RF, which is too close to the Chinstrap group. On the other hand between Adelie and Gentoo, the RF is better situated about half-way between the two groups. Interestingly, the RF boundaries, which can be highly nonlinear, albeit blocky, are relatively linear for this problem. }
\label{slice1}
\end{figure*}

Finally, to determine which model describes the boundaries better,
compare them with the \(A_2\) projection of the data (bottom row of
Figure \ref{slice1}). Gentoo is more separated from the other two in
this projection, and one can imagine that the trees in the forest has
greedily grasped any one of many places to make a split to separate the
group. It might be argued though the that RF boundary is cut too close
to the Chinstrap species, and might lead to some unnecessary
misclassification with new data. The LDA boundary is better placed for
all species.

This example can be reproduced with the code in
\texttt{penguins\_slicing\_through\_center.nb} and a run through is
shown in the first part of the video at
\url{https://vimeo.com/747590472}.

@@

We start by a comparison of the two models and the data distribution
shifting to \(\mathbf{c}^{+}\), thus the slice is localized towards high
values of bill depth in top row of Figure \ref{slice1p}.
\textcolor{blue}{The star glyph at right indicates the slice center value, which shows that it is higher on `bd` and at the middle value for the other variables. This slice center guide is as proposed in @condviz2, and more details can be found in the Appendix.}
We can see that all three slices (the two models and the data) contain
almost no points from the third class (Gentoo), and that the decision
boundary between the two models is very similar.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.95\textwidth]{figures/slice1_p.pdf}}
%\centerline{\includegraphics[width=0.32\textwidth]{figures/slice1_p_rf.png}
%\includegraphics[width=0.32\textwidth]{figures/slice1_p_lda.png}
%\includegraphics[width=0.32\textwidth]{figures/slice1_p_data.png}}
\caption{Shifting the slice center allows for exploring the boundary in different subspaces. Here the center has been shifted along the axis of `bd` (icon display at right represents slice center position). (Top row) Center is at large values of `bd`, showing a region with no Gentoo. The two models have similar boundaries except that the LDA is more linear. (Bottom row) Center is at small values of `bd`, reveals a region where the models differ substantially. The RF model carves some area for Adelie, but LDA would only predict to be Gentoo or Chinstrap here. LDA better matches the observed data in this region.}
\label{slice1p}
\end{figure*}

A more interesting comparison is found for \(\mathbf{c}^{-}\), thus the
slice localized towards low values of bill depth, shown in the bottom
row of Figure \ref{slice1p}.
\textcolor{blue}{(The star glyph at right indicates the slice center value, which shows that it is lower on `bd` and at the middle value for the other variables.)}
The RF model (left) predicts all three species within this slice, with
an interesting boundary for the third class (Gentoo). On the other hand
the LDA model (middle) predominantly predicts the third class within the
slice, this appears to be enforced through the linear structure of the
model. Looking finally at the thick slice through the data we see that
there are primarily observations from this class within the slice we can
conclude that the two models have filled in the ``empty'' space (where
we do not have any training observations) in very different ways and
according to what we might expect given the model structure.

@@

This example can be reproduced with the code in
penguins\_shifting\_slice\_center.nb and a run through is shown in the
second part of the video at \url{https://vimeo.com/747590472}.

>>>>

\begin{abstract}
This paper describes new user controls for examining high-dimensional
data using low-dimensional linear projections and slices. A user can
interactively change the contribution of a given variable to a
low-dimensional projection, which is useful for exploring the
sensitivity of structure to particular variables. The user can also
interactively shift the center of a slice, for example, to explore how
structure changes in local subspaces. The Mathematica package as well as
example notebooks are provided, which contain functions enabling the
user to experiment with these new manual controls, with one specifically
for exploring regions and boundaries produced by classification models.
The advantage of Mathematica is its linear algebra capabilities and
interactive cursor location controls. Some limited implementation has
also been made available in the R package \tt{tourr}. 
\end{abstract}

@@

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications,
\textcolor{blue}{methods which work for arbitrary dimensions have more general utility}.
A good approach is to show projections from an arbitrary dimensional
space to create dynamic data visualizations called \emph{tours}. Tours
involve views of high-dimensional data with low-dimensional projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. (Various videos of tours, e.g.
\citet{dataviewer}, \citet{gt-pp-video}, on the \citet{ASA22},
illustrate this concept.) Prior to Asimov's work, there were numerous
preparatory developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video \citep{PRIM9-video}
illustrating the capabilities is available through \textcolor{blue}{the}
video library of \citet{ASA22}.) Steering through all possible
projections is impossible, unlike Asimov's tours which allow one to
quickly see many, many different projections. After Asimov, there have
been many tour developments, which are summarized in \citet{lee2021}.

One such direction of work develops the ideas from PRIM-9, to provide
manual control of a tour. \citet{cook_manual_1997} describe controls for
1D (or 2D) projections, respectively in a 2D (or 3D) manipulation space,
allowing the user to select any variable axis, and rotate it into, out
of, or around the projection through horizontal, vertical, oblique,
radial or angular changes in value. \citet{spyrison_spinifex_2020}
refined this algorithm and implemented it to generate radial tour
animation sequences.

Manual controls are especially useful for assessing the sensitivity of
structure to particular elements of the projection. There are many
places where it is useful. In exploratory data analysis, where one sees
clusters in a projection, one may ask whether some variables can be
removed from the projection without affecting the clustering. For
interpreting models, one can reduce or increase a variable's
contribution, to examine the variable importance. Having the user
interact with a projection is extremely valuable for understanding
high-dimensional data. However, these algorithms have two problems: (1)
the pre-processing of creating a manipulation space overly complicates
the algorithm, and (2) extending to higher dimensional control is
difficult.

Another potentially useful manual control is to allow the user to choose
the position of the center of a slice. The slice tour was introduced in
\citet{slicetour}. It operates by converting the projection plane into a
slice, by removing or de-emphasizing points that are further than a
fixed orthogonal distance from the plane. The projection plane is
usually thought of as passing through the center of the data. Manual
control would allow the user to change the position of the center point,
by shifting it along a coordinate axis, while keeping the orientation of
the projection plane fixed. The purpose would be to explore how or if
the shape of the data, in the space orthogonal to the projection,
changes as one gets away from the center. It would also allow the user
to interactively decide on the thickness of the slice.

This paper explains the new manual controls for projection and slice
tours. The next section describes the new algorithm for manual control,
for both projections and slices. The use of these methods is illustrated
to compare and contrast boundaries constructed by different classifiers.
The software section describes a Mathematica \citep{Mathematica} package
that is used for the application and describes the interactive
environment that would be desirable within R \citep{rref} as new
technology becomes available. The paper is accompanied by an appendix
with more details and adjustments to the manual controls, and three
Mathematica notebooks that can be used to reproduce the applications.

@@

The methods for updating component values in \citet{cook_manual_1997}
(and utilized in \citet{spyrison_spinifex_2020}) are prescribed
primarily for a 2D projection, to take advantage of (then) newly
developed 3D trackball controls made available for computer gaming. The
first step was to construct a 3D manipulation space from a 2D
projection. In this space, the coefficient of the controlled variable
ranges between -1 and 1. Movements of a cursor are recorded and
converted into changes in the values of \(V_m\) thus changing the
displayed 2D projection. The algorithm also provided constraints to
horizontal, vertical, radial, or angular motions only. The construction
of the manipulation space overly complicates the manual controls,
especially when considering possible techniques that will apply to
arbitrary \(d\).

@@

The new approach emerged from experiments on the tour using linear
algebra capabilities, and a relatively new interactive graphics
interface, available in Mathematica. The components corresponding to
\(V_m\) are directly controlled by cursor movement, which updates row
\(m\) of \(A\). The updated matrix is then orthonormalized.

@@

This algorithm will produce the changes to a projection as illustrated
in Figure \ref{fig:manualsequence}. The controlled variable, \(V_m\),
corresponds to the dark line, and sequential changes to row \(m\) of
\(A\) can be seen to roughly follow a specified position (dot). Changes
in the other components happen as a result of the orthonormalization but
are uncontrolled.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/manualsequence-1} \caption{Sequence of projections where the contribution of one variable is controlled (dark) is changed using unconstrained orthonormalization. The dot indicates the chosen values for the controlled variable, $V_m$. It can be seen that the actual axis does not precisely match the chosen position, but it is close.}\label{fig:manualsequence}
\end{figure}

@@

The problem with the new simple method is that it is not faithful to the
precise values for \(V_m\) because the orthonormalization will change
them. Even though these changes are for the most part imperceptible, one
may wish to avoid them and there are numerous ways that this can be
enforced, a few are detailed in the Appendix. These primarily differ in
how the remaining variables are adjusted during orthonormalization.

@@

To better explore the space we combine the manual controls for the
projection with manual controls for slicing. A slice is a section of the
data that is defined by a projection, a center point that is anchoring
it in the high-dimensional space, and the slice thickness \(h\)
\citep{slicetour}. A data point is inside the slice if its orthogonal
distance from the projection plane (passing through the center point) is
below the thickness \(h\). This orthogonal distance is computed in terms
of the component that is normal on the projection plane. For
\(\mathbf{x}_i\) a \(p\) dimensional data point and \(\mathbf{c}\) the
center point (in the same \(p\) dimensional space) we compute the
orthogonal distance as \begin{equation}
v_i^2 = ||\mathbf{x}_i' - \mathbf{c}'||^2 = \mathbf{x}_i'^2 + \mathbf{c}'^2 - 2 \mathbf{x}_i'\cdot\mathbf{c}' ,
\label{eq:slice}
\end{equation} with
\(\mathbf{c}' = \mathbf{c} - (\mathbf{c}\cdot \mathbf{a}_1) \mathbf{a}_1 - (\mathbf{c}\cdot \mathbf{a}_2 )\mathbf{a}_2\),
\(\mathbf{x}_i' = \mathbf{x}_i - (\mathbf{x}_i\cdot \mathbf{a}_1) \mathbf{a}_1 - (\mathbf{x}_i\cdot \mathbf{a}_2) \mathbf{a}_2\)
and \(\mathbf{a}_k, k=1,2 (=d)\) denoting the columns of the projection
matrix, \(\mathbf{A}=(\mathbf{a}_1, \mathbf{a}_2)\).

@@

In addition, it is also useful to interactively change the slice
thickness \(h\) (also called the slice radius), in particular, to find
the preferred value for exploring the input data. For guidance the
estimates of the number of points inside the slice as a function of the
original sample size \(N\) and the number of dimensions \(p\) from
\citet{sectionpursuit} can be used: in case of a uniform distribution
inside a sphere of radius \(R\) a slice with thickness \(h\) will
contain \(N_S\) points, with \begin{equation}
N_S(h, p, R, N) = \frac{N}{2} \left(\frac{h}{R}\right)^{p-2} \left(p - (p-2)\left(\frac{h}{R}\right)^{2}\right).
\label{eq:count}
\end{equation}

@@

In addition, we also want to be able to look at slices of the data and
select slicing parameters interactively. Here we have implemented a
switch to change between projection and slice view, a slider that
adjusts the slice thickness, \(h\) (named \texttt{height} in the
Mathematica interface), and a numeric input to specify the center point
\(c\) explicitly. The display will update based on these input values,
in particular, the view will jump to a new slice when \(c\) is changed.

\textcolor{blue}{Within R, the natural implementation for an interactive interface would use}
\texttt{shiny} \citep{shinym} with \texttt{plotly} \citep{plotly}
drawing. However, the \texttt{plotly} package is designed to add to
\texttt{ggplot2} graphics, not to be a programmable interactive graphics
system. It lacks the capability to implement the primary tool needed for
a manual tour, which is the tracking of the cursor position in small
increments. In theory, the historical \texttt{locator()} function of
base R should provide the necessary capability, but its behavior is not
consistent between graphical devices. The cursor position control is
possible with \texttt{rgl} \citep{rgl} but this package is focused on
3D, not the arbitrary dimension of tours. The level of interactivity
would have been possible with the \texttt{cranvas} system
\citep{cranvas} or \texttt{rJava} \citep{rJava} but neither has a
graphics toolkit that currently operates across platforms. What we have
chosen to do is to use the interactivity of Mathematica to experiment,
and to implement the resulting ideas as an animation in the
\texttt{tourr} \citep{tourr} package. There is hope that better
interactivity will be available for R soon, perhaps in javascript
\citep[e.g.][\citet{detourr}]{langevitour}, which might allow for the
experimentation to develop new techniques.

@@

Changing the slice center manually can be accomplished with the new
function, \texttt{manual\_slice()}. This changes the value of the center
point (\(\mathbf{c}\)) of the slice, along a selected variable axis. It
slices from the center of the data out to one side, back to the center,
out to the opposite side, and then back to the center.

@@

\caption{Scatterplot matrix of the (standardized) penguins data. The three species are reasonably different in size, with Gentoo distinguished from the other two on body depth relative to flipper length and body mass.}\label{fig:penguins-scatmat}
\end{figure}

To illustrate the usefulness of the manual controls we use the 4D
penguins data \citep{penguins} and look at classification models
following \citet{sam11271}. We will show how classification boundaries
can be explored and better understood on projections and slices through
4D space. Figure \ref{fig:penguins-scatmat} shows a scatterplot matrix
of this data. There are four variables
(\texttt{bl\ =\ bill\_length\_mm,\ bd\ =\ bill\_depth\_mm,\ fl\ =\ flipper\_length\_mm,\ bm\ =\ body\_mass\_g})
measuring the size of the penguins from three species (Adelie,
Chinstrap, and Gentoo). The scatterplot matrix shows that the three
species appear to be likely separable and that at least the Gentoo can
be distinguished from the other two species when \texttt{bd} is paired
with \texttt{fl} or \texttt{bm}. The steps for exploring boundaries in
this example are as follows:

@@

We start by exploring the projections of the model prediction. Figure
\ref{proj1} summarizes the process. Rotating manually the view, we can
visualize the location of predictions for each of the three species, and
also get a sense of the difference between the two models. To illustrate
this difference, we have manually rotated the projection for the RF
model (left plot) to identify a view that shows the non-linear but
block-type structure that is typical of this type of model. This
particular projection (\(A_1\)) is exported so that it can later be used
to show the LDA model (middle plot) and the actual data (right plot).
What can be seen is the linearity of the LDA model, where the boundaries
are linear and oblique to the variable axes. And, interestingly, this
particular projection of the original data shows very distinct clusters
of the three species. That means the obscuring of the boundaries between
groups for both of the models is driven by what is happening in the
orthogonal space to the plane of the selected projection.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.95\textwidth]{figures/proj1.pdf}}
%\centerline{\includegraphics[width=0.32\textwidth]{figures/proj1_rf.png}
%\includegraphics[width=0.32\textwidth]{figures/proj1_lda.png}
%\includegraphics[width=0.32\textwidth]{figures/proj1_data.png}}
\caption{A projection identified using the manual tour because it shows differences in classification boundaries induced by different models. An RF model produces blocky, rectangular, boundaries, while the LDA model produces oblique linear boundaries. For comparison, in this projection, the data shows the three groups are almost perfectly separable.}
\label{proj1}
\end{figure*}

This example can be reproduced with the code in
penguins\_exploring\_manually.nb and a run-through is shown in the video
at \url{https://vimeo.com/747585410}.

@@

We now continue the investigation by slicing orthogonally to the
projection \(A_1\). For both models we look at a thin (\(h=0.5\)) slice
through the center, \(\mathbf{c}^0 = (0,0,0,0)\). At first, we explore
how changing the projection away from \(A_1\) can help with
understanding the boundary better. For our example, notice that \(A_1\)
does not contain any contribution from the second variable
(\texttt{bd}), so we will first rotate this variable into the view.

@@

The RF is almost straight in this view, as something we might expect
from a tree model where splitting occurs on single variables only.
However, it is straight in a combination of the second and third
variables (\texttt{bd} and \texttt{fl}). By examining the scatterplot
matrix, we can understand how this boundary was built: on each of
\texttt{bd} and \texttt{fl}, it is possible to cut on a value where most
of the Gentoo penguins are different from the other two species for any
tree, and likely only one is needed. Thus the forest construction is
providing a sample of trees that use one variable or the other variable
to split, producing the blocky boundary on the combination of the two.

In contrast, the LDA boundary has an oblique split between the two and
roughly divides the space into three similarly sized areas. It is almost
like a textbook illustration of how LDA works for two variables (2D)
where by assuming the clusters have equal variance-covariance, it places
a boundary halfway between the group means.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.95\textwidth]{figures/proj2.pdf}}
%\centerline{\includegraphics[width=0.45\textwidth]{figures/slice1_rf.png}
%\includegraphics[width=0.45\textwidth]{figures/slice1_lda.png}}
%\centerline{\includegraphics[width=0.45\textwidth]{figures/slice2_rf.png}
%\includegraphics[width=0.45\textwidth]{figures/slice2_lda.png}}
\caption{Comparing slices from two different projections $A_1$ (top row) and $A_2$ (bottom row), to compare and contrast classification boundaries induced by two models RF and LDA. A thin slice through the center of the data is shown. The plot of the data shows that the Gentoo group (light grey/green) is less separated in $A_1$, and the rotation to $A_2$ produces a bigger gap. The boundaries between groups are clear in $A_2$ also. The LDA boundary is more oblique between Gentoo and Chinstrap than the RF, which is too close to the Chinstrap group. On the other hand between Adelie and Gentoo, the RF is better situated about halfway between the two groups. Interestingly, the RF boundaries, which can be highly nonlinear, albeit blocky, are relatively linear for this problem. }
\label{slice1}
\end{figure*}

Finally, to determine which model describes the boundaries better,
compare them with the \(A_2\) projection of the data (bottom row of
Figure \ref{slice1}). Gentoo is more separated from the other two in
this projection, and one can imagine that the trees in the forest have
greedily grasped any one of many places to make a split to separate the
group. It might be argued though the that RF boundary is cut too close
to the Chinstrap species, and might lead to some unnecessary
misclassification with new data. The LDA boundary is better placed for
all species.

This example can be reproduced with the code in
\texttt{penguins\_slicing\_through\_center.nb} and a run-through is
shown in the first part of the video at
\url{https://vimeo.com/747590472}.

@@

We start with a comparison of the two models and the data distribution
shifting to \(\mathbf{c}^{+}\), thus the slice is localized towards high
values of bill depth in the top row of Figure \ref{slice1p}.
\textcolor{blue}{The star glyph at right indicates the slice center value, which shows that it is higher on `bd` and at the middle value for the other variables. This slice center guide is as proposed in @condviz2, and more details can be found in the Appendix.}
We can see that all three slices (the two models and the data) contain
almost no points from the third class (Gentoo) and that the decision
boundary between the two models is very similar.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.95\textwidth]{figures/slice1_p.pdf}}
%\centerline{\includegraphics[width=0.32\textwidth]{figures/slice1_p_rf.png}
%\includegraphics[width=0.32\textwidth]{figures/slice1_p_lda.png}
%\includegraphics[width=0.32\textwidth]{figures/slice1_p_data.png}}
\caption{Shifting the slice center allows for exploring the boundary in different subspaces. Here the center has been shifted along the axis of `bd` (icon display at right represents slice center position). (Top row) The center is at large values of `bd`, showing a region with no Gentoo. The two models have similar boundaries except that the LDA is more linear. (Bottom row) The center is at small values of `bd`, revealing a region where the models differ substantially. The RF model carves some area for Adelie, but LDA would only predict to be Gentoo or Chinstrap here. LDA better matches the observed data in this region.}
\label{slice1p}
\end{figure*}

A more interesting comparison is found for \(\mathbf{c}^{-}\), thus the
slice localized towards low values of bill depth, shown in the bottom
row of Figure \ref{slice1p}.
\textcolor{blue}{(The star glyph at right indicates the slice center value, which shows that it is lower on `bd` and at the middle value for the other variables.)}
The RF model (left) predicts all three species within this slice, with
an interesting boundary for the third class (Gentoo). On the other hand,
the LDA model (middle) predominantly predicts the third class within the
slice, this appears to be enforced through the linear structure of the
model. Looking finally at the thick slice through the data we see that
there are primarily observations from this class within the slice we can
conclude that the two models have filled in the ``empty'' space (where
we do not have any training observations) in very different ways and
according to what we might expect given the model structure.

@@

This example can be reproduced with the code in
penguins\_shifting\_slice\_center.nb and a run-through is shown in the
second part of the video at \url{https://vimeo.com/747590472}.
